---
title: "PCA Projection and Missing Data"
output:
  html_document:
    df_print: paged
editor_options:
  chunk_output_type: console
---

# PCA Projection and Missing Data

### Todo
* make better section headings
* add better descriptions
* profit

### Load dataset and dependencies

Here we're loading a genotype matrix with 159 individuals and ~30k SNPs. We selected three individuals from each Human Origins population.

```{r}
# install dependencies
packages <- c("tidyverse", "cowplot", "softImpute", "missMethods", "norm", "mvtnorm", "ggrepel", "plotly", "magrittr")
Map(function(x) { install.packages(x) }, packages[!packages %in% utils::installed.packages()])

library(magrittr)
library(ggplot2)
library(plotly)

knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, fig.width = 8, fig.height = 6)


#setwd('~/Documents/exp_dat_reading_group_2021/session_4/')
source('helper_functions.R')

# load data
geno_matrix <- scan("geno_matrix_three.txt", what = "character") %>%
  strsplit("") %>%
  do.call(rbind, .) %>%
  apply(., 2, as.numeric)

context_info <- readr::read_csv("context_info_three.csv")

###### 

```

### Global PCAs

Here we plot two PCAs - the first uses all 159 individuals. For the second plot, we remove 9 individuals (plotted with red dots and labels), and then project those individuals onto a PCA generated from the remaining 150 individuals. You can see that there is some shrinkage towards the origin.

```{r}
prcomp(geno_matrix) %>% 
  tidy_pca_output(context = context_info) %>%
  dplyr::mutate(projected = 
                  dplyr::case_when(Group_Name %in% c('Papuan', 'Russian', 'Italian_North') ~ 'projected',
                                   TRUE ~ 'pca')) %>%
  plot_tidy_pca_simple(use.labels = 'projected') + ggtitle('PCA with all individuals')

plot_tidy_pca_simple(project_downsampled_inds(geno_matrix, destruction_level = 0.0, 
                                              context = context_info,
                                              drop_groups = c('Papuan', 'Russian', 'Italian_North')),
                     use.labels = 'projected') + ggtitle('PCA with Papuan, Russian, Italian_North projected')

```

### Downsample data from projected individuals

First compute the static "background" PCA.

```{r}
pca_static <- 
  project_downsampled_inds(geno_matrix, destruction_level = 0.0,
                           drop_groups = c('Papuan', 'Russian'), ret.pca_rest = T)

```

#### Projections with three levels of downsampling

```{r}
pnf_tidy_drop_ind <- project_downsampled_inds(geno_matrix, destruction_level = 0.0,
                                              drop_groups = c('Papuan', 'Russian'), 
                                              pca_rest = pca_static)
plot_tidy_pca_simple(pnf_tidy_drop_ind, use.labels = 'projected') +
  ggtitle('Projecting Papuan + Russian', subtitle = '0% missing')

pnf_tidy_drop_ind <- project_downsampled_inds(geno_matrix, destruction_level = 0.5,
                                              drop_groups = c('Papuan', 'Russian'), 
                                              pca_rest = pca_static)
plot_tidy_pca_simple(pnf_tidy_drop_ind, use.labels = 'projected') +
  ggtitle('Projecting Papuan + Russian', subtitle = '50% missing')

pnf_tidy_drop_ind <- project_downsampled_inds(geno_matrix, destruction_level = 0.9, 
                                              drop_groups = c('Papuan', 'Russian'), 
                                              pca_rest = pca_static)
plot_tidy_pca_simple(pnf_tidy_drop_ind, use.labels = 'projected') +
  ggtitle('Projecting Papuan + Russian', subtitle = '90% missing')


```

#### Now try many different downsample values, to see the effects of extremely large amounts of missing data.

```{r}

## static PCA, to be used in each projection
pca_static <- 
  project_downsampled_inds(geno_matrix, destruction_level = 0.0,
                           drop_groups = c('Papuan', 'Russian'), ret.pca_rest = T)

## projection w/ no missing data
pnf_tidy_drop_ind.mult <- 
  project_downsampled_inds(geno_matrix, destruction_level = 0.0,
                           drop_groups = c('Papuan', 'Russian'), pca_rest = pca_static)
pnf_tidy_drop_ind.mult <- pnf_tidy_drop_ind.mult %>% dplyr::mutate(iter = 0)

## projections with progressively more missing data
cat('Downsampling:')
iter = 1
for (ds in c(seq(.5, .9, .1), seq(.91, .999, .003))) {
  cat(' ', ds)
  tmp1 <- project_downsampled_inds(geno_matrix, ds,
                                   drop_groups = c('Papuan', 'Russian'), 
                                   pca_rest = pca_static)
  tmp1 <- tmp1 %>% dplyr::mutate(iter = iter)
  pnf_tidy_drop_ind.mult <- rbind(pnf_tidy_drop_ind.mult, tmp1)
  iter = iter + 1
}
cat('Done.\n')

```

Plot this with an animation, where you can scan through the amount of data removed.

```{r}
ggplotly(plot_tidy_pca_simple(pnf_tidy_drop_ind.mult, use.labels = 'projected'))
```


### Try to get better boundaries

Here we do many iterations of the same amount of downsampling, to better observe the range of effects.


```{r}
## static PCA, to be used in each projection
pca_static <- 
  project_downsampled_inds(geno_matrix, destruction_level = 0.0, context = context_info,
                           drop_groups = c('Papuan', 'Russian', 'Italian_North'), ret.pca_rest = T)

## projection w/ no missing data
pnf_tidy_drop_ind.mult.og <- 
  project_downsampled_inds(geno_matrix, destruction_level = 0.0, context = context_info,
                           drop_groups = c('Papuan', 'Russian', 'Italian_North'), 
                           pca_rest = pca_static)
pnf_tidy_drop_ind.mult <- pnf_tidy_drop_ind.mult.og %>% dplyr::mutate(iter = 0)

## projections with progressively more missing data
cat('Downsampling:')
iter = 1
for (ds in c(seq(.5, .9, .1), .95, .96, .97, .98, .99)) {
  cat(' ', ds)
  pnf_tidy_drop_ind.mult <- 
    rbind(pnf_tidy_drop_ind.mult,
          pnf_tidy_drop_ind.mult.og %>% dplyr::mutate(downsample = ds, iter = iter))
  
  for (x in seq(10)) {
    tmp1 <- project_downsampled_inds(geno_matrix, ds, context = context_info,
                                     drop_groups = c('Papuan', 'Russian', 'Italian_North'), 
                                     pca_rest = pca_static)
    tmp1 <- tmp1 %>% dplyr::filter(projected == 'projected')
    tmp1 <- tmp1 %>% dplyr::mutate(iter = iter)
    pnf_tidy_drop_ind.mult <- rbind(pnf_tidy_drop_ind.mult, tmp1)
  }
  iter = iter + 1
}
cat('Done.\n')

```

```{r}
ggplotly(plot_tidy_pca_simple(pnf_tidy_drop_ind.mult, use.labels = 'projected'))

```



### non-African PCA

It's possible that our dataset is not very susceptible to downsampling, due to the large amounts of variation present in the populations. Here we subset to just the non-African samples, and repeat some of the above experiments.

#### Basic non-African PCA (with and without projection)

```{r}

## keep mostly non-Africans and non-Americas
drop_ind <- which(context_info$Makro_Region %in% c('Africa', 'Americas') 
                  | is.na(context_info$Makro_Region) 
                  | context_info$Group_Name %in% c('Papuan','Nasioi','Tujia','','','',''))

# ## keep mostly West and Central Eurasian
# drop_ind <- which(context_info$Makro_Region %in% c('Africa', 'Americas', 'East Asia') 
#                   | is.na(context_info$Makro_Region) 
#                   | context_info$Group_Name %in% c('Papuan','Nasioi','Tujia','Uyghur','Hazara','Yakut',''))

## get data for rest
geno_matrix_non_afr <- geno_matrix[-drop_ind,]
dim(geno_matrix_non_afr)

## get context for rest
context_info_non_afr <- context_info[-drop_ind,]
dim(context_info_non_afr)
# context_info_non_afr$projected <- 'pca'

prcomp(geno_matrix_non_afr) %>% 
  tidy_pca_output(context = context_info_non_afr) %>%
  dplyr::mutate(projected = 
                  dplyr::case_when(Group_Name %in% c('Pathan', 'Russian', 'Italian_North') ~ 'projected',
                                   TRUE ~ 'pca')) %>%
  plot_tidy_pca_simple(use.labels = 'projected') + ggtitle('PCA with all [Eurasian] individuals')

plot_tidy_pca_simple(project_downsampled_inds(geno_matrix_non_afr, destruction_level = 0.0,
                                              context = context_info_non_afr,
                                              drop_groups = c('Pathan', 'Russian', 'Italian_North')),
                     use.labels = 'projected') + 
  ggtitle('PCA with Pathan, Russian, Italian_North projected')


```


#### Try many levels of downsampling

```{r}
## static PCA, to be used in each projection
pca_static.small <- 
  project_downsampled_inds(geno_matrix_non_afr, destruction_level = 0.0, context = context_info_non_afr,
                           drop_groups = c('Pathan', 'Russian', 'Italian_North'), ret.pca_rest = T)

## projection w/ no missing data
pnf_tidy_drop_ind.mult.small <- 
  project_downsampled_inds(geno_matrix_non_afr, destruction_level = 0.0, context = context_info_non_afr,
                           drop_groups = c('Pathan', 'Russian', 'Italian_North'), 
                           pca_rest = pca_static.small)
pnf_tidy_drop_ind.mult.small <- pnf_tidy_drop_ind.mult.small %>% dplyr::mutate(iter = 0)

## projections with progressively more missing data
cat('Downsampling:')
iter = 1
for (ds in c(seq(.5, .9, .1), seq(.91, .999, .003))) {
  cat(' ', ds)
  tmp1 <- project_downsampled_inds(geno_matrix_non_afr, ds, context = context_info_non_afr,
                                   drop_groups = c('Pathan', 'Russian', 'Italian_North'), 
                                   pca_rest = pca_static.small)
  tmp1 <- tmp1 %>% dplyr::mutate(iter = iter)
  pnf_tidy_drop_ind.mult.small <- rbind(pnf_tidy_drop_ind.mult.small, tmp1)
  iter = iter + 1
}
cat('Done.\n')

```

```{r}
ggplotly(plot_tidy_pca_simple(pnf_tidy_drop_ind.mult.small, use.labels = 'projected'))

```




## Debug PCA shrinkage

Ignore this

```{r}
tmp.pca_static.small.nodrops <- prcomp(geno_matrix_non_afr)

## static PCA, to be used in each projection
tmp.pca_static.small.drops <- 
  project_downsampled_inds(geno_matrix_non_afr, destruction_level = 0.0, context = context_info_non_afr,
                           drop_groups = c('Pathan', 'Russian', 'Italian_North'), ret.pca_rest = T)

plot(tmp.pca_static.small.nodrops$rotation[, 1], tmp.pca_static.small.drops$rotation[, 1])
plot(tmp.pca_static.small.nodrops$rotation[, 2], tmp.pca_static.small.drops$rotation[, 2])

cor(tmp.pca_static.small.nodrops$rotation[, 1], tmp.pca_static.small.drops$rotation[, 1])
cor(tmp.pca_static.small.nodrops$rotation[, 2], tmp.pca_static.small.drops$rotation[, 2])

```



Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Cmd+Option+I*.

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Cmd+Shift+K* to preview the HTML file). 

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.

