---
title: "Lesson 1: Fill in missing data"
author: "Clemens Schmid"
date: "10/13/2021"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Missing data

Missing information is one of the core challenges for working with ancient DNA data. Here we want to explore its effect on PCA.

### Dependencies

This analysis relies on a number of R packages. You can install them with the following command:

```{r,results=F}
packages <- c("tidyverse", "cowplot", "softImpute", "missMethods", "norm", "mvtnorm", "ggrepel")
Map(function(x) { install.packages(x) }, packages[!packages %in% utils::installed.packages()])
```

```{r}
library(magrittr)
library(ggplot2)
```

### Dataset

We prepared a simplified genotype dataset, which should allow you to explore the effect of missingness in the (familiar?) environment of world-wide population history. It contains one individual each from about 50 populations from around the world. The `context_info` table holds some minimal context information.

```{r}
context_info <- readr::read_csv("context_info_one.csv")
context_info
```

```{r,warning=FALSE}
world <- map_data("world")
ggplot() +
  geom_map(data = world, map = world, aes(long, lat, map_id = region)) +
  geom_point(data = context_info, aes(Longitude, Latitude, colour = Makro_Region), size = 3) +
  coord_fixed()
```

For each individual we extracted the same <50000 non-missing SNPs from the HumanOrigins SNP array (Patterson 2012). They are available in a large `geno_matrix` text filem wehich can be loaded with

```{r}
geno_matrix <- scan("geno_matrix_one.txt", what = "character") %>%
  strsplit("") %>%
  do.call(rbind, .) %>%
  apply(., 2, as.numeric)
```

This large matrix fills contains more than 2 million entries and requires about 20mb of memory. Nevertheless we can still conveniently and interactively work with it in R. It features `r nrow(geno_matrix)` rows, one for each individual, and one column for each of the `r ncol(geno_matrix)` SNPs. Each cell has either the value 0, 1 or 2 and -- although this is technically discrete data -- we'll treat it as continuously scaled.

```{r}
#table(geno_matrix)
```

### Running PCA

To explore this complex dataset we can run PCA on it, e.g. with `stats::prcomp()`:

```{r}
pca <- prcomp(geno_matrix)
```

The output object of this function is a complex beast.

```{r}
str(pca)
```

Though most of the time we're only interested in three fields for concrete applications:

- `$sdev`: standard deviations of the principal components
- `$rotation`: the matrix of variable loadings
- `$x`: the value of the rotated data, so our input observations coordinates in PCA space

From the standard deviations (`$sdev`) we can calculate the proportions of variance covered by the individual principal components:

```{r}
pov <- pca$sdev^2/sum(pca$sdev^2)
pov_df <- tibble::tibble(pc = as.factor(1:10), pov = pov[1:10])
pov_df %>% ggplot() + geom_bar(aes(pc, pov), stat = "identity")
```

The variable loadings (`$rotation`) can be explored individually (though it's a bit tedious for tens of thousands of SNPs) or used for projection (see lesson 2).

The output coordinates (`$x`) finally are most immediately useful to plot a "genetic map". 

```{r}
pca$x %>% 
  as.data.frame() %>%
  ggplot() +
  geom_point(aes(PC1, PC2))
```

This alone is of doubtful usefulness, as we can't really interpret a random, unlabelled point cloud. We have to add meaningful context information. Fortunately we already have this available in the `context_info` object` and can combine the two.

```{r}
pca_tidy_obs <- pca$x %>%
  tibble::as_tibble() %>%
  dplyr::bind_cols(context_info)

pca_tidy_obs[c(1,2,(ncol(pca_tidy_obs)-5):ncol(pca_tidy_obs))]
```

With this context information, the PCA plot can be made a lot more informative and useful.

```{r}
pca_tidy_obs %>% 
  ggplot() +
  geom_point(aes(PC1, PC2, color = Makro_Region))
```

[Let's discuss this plot!]

We automated the described data preparation and plotting operations with a number of helper functions in a supplementary script file. They also include a plotting function, that adds marginal kernel density plots to our default figure.

```{r}
source("helper_functions.R")

prcomp(geno_matrix) %>%
  tidy_pca_output() %>%
  plot_tidy_pca_density(ggrepel::geom_text_repel)
```

[What do the axis density plots tell us?]

### Creating missing data

To explore the effect of missing data on the PCA plot, we can artificially modify our currently complete dataset. Figuratively we can "shoot holes" into it. Here's a simple  matrix:

```{r}
x <- matrix(1:9, nrow = 3, ncol = 3, byrow = T)
x
```

We can poke holes into it with a function like this:

```{r}
# shoot_holes()
prop <- 0.3 # desired percentage of missingness
nr_holes <- round(prop * prod(dim(x)))
holes <- sample(seq_along(x), size = nr_holes)
x_Emmentaler <- x
x_Emmentaler[holes] <- NA
x_Emmentaler
```

This implementation has the issue, that it can accidentally wipe out entire columns. We want to avoid that for this experiment, so here's an alternative implementation that removes the same proportion of values from each column.

```{r}
# shoot_holes_column_wise()
prop <- 0.3 # desired percentage of missingness
nr_holes_per_column <- round((prop * prod(dim(x)))/ncol(x))
x_Tilsit <- apply(x, 2, function(y) {
  holes <- sample(seq_along(y), size = nr_holes_per_column)
  y[holes] <- NA
  y
})
x_Tilsit
```

Equipped like this we can start to poke holes into our genotype data and run PCA on it.

```{r}
geno_perforated <- shoot_holes_column_wise(geno_matrix, 0.3)
prcomp(geno_perforated)
```

This yields a nasty error message. `stats::prcomp()` can not run with perforated input data. So we're forced to patch the holes.

### Filling missing data

There are multiple algorithms to consider here - as discussed in our meeting last week. 

#### Mean imputation

Mean imputation fills the missing values with the means of some observed values. We can write a column-wise mean imputation like this:

```{r}
patch_holes_mean <- function(x) {
  apply(x, 2, function(y) { 
    y[is.na(y)] <- mean(y, na.rm = T)
    y
  })
}
```

... and then run it on our perforated dataset:

```{r}
set.seed(123)

geno_matrix %>%
  shoot_holes_column_wise(0.5) %>%
  patch_holes_mean() %>%
  prcomp(geno_matrix) %>%
  tidy_pca_output() %>%
  plot_tidy_pca_simple(ggrepel::geom_text_repel, max.overlaps = Inf)
```

[Compare the plot to the one without imputed data!]

The R package `missingMethods` provides a number of imputation algorithms, including the following three for mean imputation: `impute_mean()`, `impute_median()`, and `impute_mode`. The first one should behave identically to our own implementation `patch_holes_mean()`. We can compare their effects:

```{r}
geno_matrix %>% explore_filling_method(missMethods::impute_mean, 0.5)
geno_matrix %>% explore_filling_method(missMethods::impute_median, 0.5)
geno_matrix %>% explore_filling_method(missMethods::impute_mode, 0.5)
```



