---
title: "R Notebook"
output: html_notebook
editor_options: 
  chunk_output_type: console
---



```{r}
# install dependencies
packages <- c("tidyverse", "cowplot", "softImpute", "missMethods", "norm", "mvtnorm", "ggrepel", "plotly", "magrittr")
Map(function(x) { install.packages(x) }, packages[!packages %in% utils::installed.packages()])

library(magrittr)
library(ggplot2)
library(plotly)

setwd('~/Documents/exp_dat_reading_group_2021/session_4/')
source('helper_functions.R')

# load data
geno_matrix <- scan("geno_matrix_three.txt", what = "character") %>%
  strsplit("") %>%
  do.call(rbind, .) %>%
  apply(., 2, as.numeric)

context_info <- readr::read_csv("context_info_three.csv")

###### 

```


```{r}
prcomp(geno_matrix) %>% 
  tidy_pca_output(context = context_info) %>%
  dplyr::mutate(projected = 
                  dplyr::case_when(Group_Name %in% c('Papuan', 'Russian', 'Italian_North') ~ 'projected',
                                   TRUE ~ 'pca')) %>%
  plot_tidy_pca_simple(use.labels = 'all') + ggtitle('PCA with all [Eurasian] individuals')

plot_tidy_pca_simple(project_downsampled_inds(geno_matrix, destruction_level = 0.0, 
                                              context = context_info,
                                              drop_groups = c('Papuan', 'Russian', 'Italian_North')),
                     use.labels = 'all') + ggtitle('PCA with Papuan, Russian, Italian_North projected')

```


```{r}
pca_static <- 
  project_downsampled_inds(geno_matrix, destruction_level = 0.0,
                           drop_groups = c('Papuan', 'Russian'), ret.pca_rest = T)

```


```{r}
pnf_tidy_drop_ind <- project_downsampled_inds(geno_matrix, destruction_level = 0.0,
                                              drop_groups = c('Papuan', 'Russian'), 
                                              pca_rest = pca_static)
plot_tidy_pca_simple(pnf_tidy_drop_ind, use.labels = 'projected') +
  ggtitle('Projecting Papuan + Russian', subtitle = '0% missing')

pnf_tidy_drop_ind <- project_downsampled_inds(geno_matrix, destruction_level = 0.5,
                                              drop_groups = c('Papuan', 'Russian'), 
                                              pca_rest = pca_static)
plot_tidy_pca_simple(pnf_tidy_drop_ind, use.labels = 'projected') +
  ggtitle('Projecting Papuan + Russian', subtitle = '50% missing')

pnf_tidy_drop_ind <- project_downsampled_inds(geno_matrix, destruction_level = 0.9, 
                                              drop_groups = c('Papuan', 'Russian'), 
                                              pca_rest = pca_static)
plot_tidy_pca_simple(pnf_tidy_drop_ind, use.labels = 'projected') +
  ggtitle('Projecting Papuan + Russian', subtitle = '90% missing')


```

Now try different levels of downsampling.

```{r}

## static PCA, to be used in each projection
pca_static <- 
  project_downsampled_inds(geno_matrix, destruction_level = 0.0,
                           drop_groups = c('Papuan', 'Russian'), ret.pca_rest = T)

## projection w/ no missing data
pnf_tidy_drop_ind.mult <- 
  project_downsampled_inds(geno_matrix, destruction_level = 0.0,
                           drop_groups = c('Papuan', 'Russian'), pca_rest = pca_static)
pnf_tidy_drop_ind.mult <- pnf_tidy_drop_ind.mult %>% dplyr::mutate(iter = 0)

## projections with progressively more missing data
cat('Downsampling:')
iter = 1
for (ds in c(seq(.5, .9, .1), seq(.91, .999, .003))) {
  cat(' ', ds)
  tmp1 <- project_downsampled_inds(geno_matrix, ds,
                                   drop_groups = c('Papuan', 'Russian'), 
                                   pca_rest = pca_static)
  tmp1 <- tmp1 %>% dplyr::mutate(iter = iter)
  pnf_tidy_drop_ind.mult <- rbind(pnf_tidy_drop_ind.mult, tmp1)
  iter = iter + 1
}
cat('Done.\n')

```

Plot this with an animation, where you can scan through the amount of data removed.

```{r}
ggplotly(plot_tidy_pca_simple(pnf_tidy_drop_ind.mult, use.labels = 'projected'))
```


### Try to get better boundaries

Lots of iterations of the same amount of downsampling


```{r}
## static PCA, to be used in each projection
pca_static <- 
  project_downsampled_inds(geno_matrix, destruction_level = 0.0, context = context_info,
                           drop_groups = c('Papuan', 'Russian', 'Italian_North'), ret.pca_rest = T)

## projection w/ no missing data
pnf_tidy_drop_ind.mult.og <- 
  project_downsampled_inds(geno_matrix, destruction_level = 0.0, context = context_info,
                           drop_groups = c('Papuan', 'Russian', 'Italian_North'), 
                           pca_rest = pca_static)
pnf_tidy_drop_ind.mult <- pnf_tidy_drop_ind.mult.og %>% dplyr::mutate(iter = 0)

## projections with progressively more missing data
cat('Downsampling:')
iter = 1
for (ds in c(seq(.5, .9, .1), .95, .96, .97, .98, .99)) {
  cat(' ', ds)
  pnf_tidy_drop_ind.mult <- 
    rbind(pnf_tidy_drop_ind.mult,
          pnf_tidy_drop_ind.mult.og %>% dplyr::mutate(downsample = ds, iter = iter))
  
  for (x in seq(10)) {
    tmp1 <- project_downsampled_inds(geno_matrix, ds, context = context_info,
                                     drop_groups = c('Papuan', 'Russian', 'Italian_North'), 
                                     pca_rest = pca_static)
    tmp1 <- tmp1 %>% dplyr::filter(projected == 'projected')
    tmp1 <- tmp1 %>% dplyr::mutate(iter = iter)
    pnf_tidy_drop_ind.mult <- rbind(pnf_tidy_drop_ind.mult, tmp1)
  }
  iter = iter + 1
}
cat('Done.\n')

```

```{r}
ggplotly(plot_tidy_pca_simple(pnf_tidy_drop_ind.mult, use.labels = 'projected'))

```



### Try with a smaller PCA

Subset to just the non-African samples.

```{r}

## keep mostly non-Africans and non-Americas
drop_ind <- which(context_info$Makro_Region %in% c('Africa', 'Americas') 
                  | is.na(context_info$Makro_Region) 
                  | context_info$Group_Name %in% c('Papuan','Nasioi','Tujia','','','',''))

# ## keep mostly West and Central Eurasian
# drop_ind <- which(context_info$Makro_Region %in% c('Africa', 'Americas', 'East Asia') 
#                   | is.na(context_info$Makro_Region) 
#                   | context_info$Group_Name %in% c('Papuan','Nasioi','Tujia','Uyghur','Hazara','Yakut',''))

## get data for rest
geno_matrix_non_afr <- geno_matrix[-drop_ind,]
dim(geno_matrix_non_afr)

## get context for rest
context_info_non_afr <- context_info[-drop_ind,]
dim(context_info_non_afr)
# context_info_non_afr$projected <- 'pca'

prcomp(geno_matrix_non_afr) %>% 
  tidy_pca_output(context = context_info_non_afr) %>%
  dplyr::mutate(projected = 
                  dplyr::case_when(Group_Name %in% c('Pathan', 'Russian', 'Italian_North') ~ 'projected',
                                   TRUE ~ 'pca')) %>%
  plot_tidy_pca_simple(use.labels = 'all') + ggtitle('PCA with all [Eurasian] individuals')

plot_tidy_pca_simple(project_downsampled_inds(geno_matrix_non_afr, destruction_level = 0.0, 
                                              context = context_info_non_afr,
                                              drop_groups = c('Pathan', 'Russian', 'Italian_North')),
                     use.labels = 'all') + ggtitle('PCA with Pathan, Russian, Italian_North projected')


```



```{r}
## static PCA, to be used in each projection
pca_static.small <- 
  project_downsampled_inds(geno_matrix_non_afr, destruction_level = 0.0, context = context_info_non_afr,
                           drop_groups = c('Pathan', 'Russian', 'Italian_North'), ret.pca_rest = T)

## projection w/ no missing data
pnf_tidy_drop_ind.mult.small <- 
  project_downsampled_inds(geno_matrix_non_afr, destruction_level = 0.0, context = context_info_non_afr,
                           drop_groups = c('Pathan', 'Russian', 'Italian_North'), 
                           pca_rest = pca_static.small)
pnf_tidy_drop_ind.mult.small <- pnf_tidy_drop_ind.mult.small %>% dplyr::mutate(iter = 0)

## projections with progressively more missing data
cat('Downsampling:')
iter = 1
for (ds in c(seq(.5, .9, .1), seq(.91, .999, .003))) {
  cat(' ', ds)
  tmp1 <- project_downsampled_inds(geno_matrix_non_afr, ds, context = context_info_non_afr,
                                   drop_groups = c('Pathan', 'Russian', 'Italian_North'), 
                                   pca_rest = pca_static.small)
  tmp1 <- tmp1 %>% dplyr::mutate(iter = iter)
  pnf_tidy_drop_ind.mult.small <- rbind(pnf_tidy_drop_ind.mult.small, tmp1)
  iter = iter + 1
}
cat('Done.\n')

```

```{r}
ggplotly(plot_tidy_pca_simple(pnf_tidy_drop_ind.mult.small, use.labels = 'projected'))

```





Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Cmd+Option+I*.

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Cmd+Shift+K* to preview the HTML file). 

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.

